---
title: "Assignment 6: GLMs (Linear Regressios, ANOVA, & t-tests)"
author: "Curtis Cha"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on generalized linear models. 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your last name into the file name (e.g., "Fay_A06_GLMs.Rmd") prior to submission.

The completed exercise is due on Monday, February 28 at 7:00 pm.

## Set up your session 
1. Set up your session. Check your working directory. Load the tidyverse, agricolae and other needed packages. Import the *raw* NTL-LTER raw data file for chemistry/physics (`NTL-LTER_Lake_ChemistryPhysics_Raw.csv`). Set date columns to date objects.

2. Build a ggplot theme and set it as your default theme.

```{r setup}
#1
library(tidyverse)
library(lubridate)
library(ggplot2)

ntl <- read.csv(
  '../Data/Raw/NTL-LTER_Lake_ChemistryPhysics_Raw.csv', 
  header = TRUE)

ntl$sampledate <- mdy(ntl$sampledate)

#2
mytheme <- theme_classic(base_size = 12) + 
  theme(axis.text = element_text(color = "black"), legend.position = "right")


```

## Simple regression
Our first research question is: Does mean lake temperature recorded during July change with depth across all lakes?

3. State the null and alternative hypotheses for this question:
> Answer:
H0: Lake depth level has no significant effect in determining mean of recorded 
lake temperatures for the month of July.
Ha: Lake depth level has a significant effect in determining mean of recorded 
lake temperatures for the month of July.


4.  Wrangle your NTL-LTER dataset with a pipe function so that the records meet the following criteria: 
 * Only dates in July. 
 * Only the columns: `lakename`, `year4`, `daynum`, `depth`, `temperature_C`
 * Only complete cases (i.e., remove NAs)

5. Visualize the relationship among the two continuous variables with a scatter plot of temperature by depth. Add a smoothed line showing the linear model, and limit temperature values from 0 to 35 °C. Make this plot look pretty and easy to read.

```{r scatterplot}
#4
ntl_july <- ntl %>% 
  mutate(month = month(ntl$sampledate)) %>%
  filter(month == 7, !is.na(ntl$temperature_C)) %>%
  select("lakename", "year4", "daynum", "depth", "temperature_C")

#5
ggplot(data = ntl_july, aes(x = depth, y = temperature_C)) +
  geom_point() + geom_smooth(method = "lm") + ylim(c(0,35)) + 
  xlab("Depth(m)") + ylab("Temperature(C)") + 
  ggtitle("Mean Recorded Temperature of Lakes (July)\n vs. Depth Level")

```


6. Interpret the figure. What does it suggest with regards to the response of temperature to depth? Do the distribution of points suggest about anything about the linearity of this trend?

> Answer: 
The scatter plot suggests that depth has a negative relationship with mean 
recorded temperature. The distribution of points suggest there may not be a 
linear relationship between the two variables sa there is high variance in 
mean temperatures at lower depts (0-5m). However at lower depths (10-15m), there
is consistent low mean temperatures. It's possible that there true relationship
between depth and temperature is not linear.

7. Perform a linear regression to test the relationship and display the results

```{r linear.regression}
#7

lm0 <- lm(data = ntl_july, temperature_C ~ depth)
sum_0 <- summary(lm0)
print(sum_0)


```


8. Interpret your model results in words. Include how much of the variability in temperature is explained by changes in depth, the degrees of freedom on which this finding is based, and the statistical significance of the result. Also mention how much temperature is predicted to change for every 1m change in depth. 

> Answer:
According to the model, for every one unit increase in depth, there is a 1.95621 
decrease in recorded mean temperature. Given the null hypothesis is true, the 
probability of observing this relationship between temperature and depth is 
significantly low (less than 0.05). Due to this low p-value, I reject the null 
hypothesis that depth has no significant effect on mean recorded lake
temperature in July. The linear model produce explains around 73.87% of the 
variability in mean recorded temperature.

---

## Multiple regression
Let's tackle a similar question from a different approach. Here, we want to explore what might the best set of predictors for lake temperature in July across the monitoring period at the North Temperate Lakes LTER. 


9. Run an AIC to determine what set of explanatory variables (year4, daynum, depth) is best suited to predict temperature.

10. Run a multiple regression on the recommended set of variables. 

```{r temperature.model}
#9

lm1 <- lm(data = ntl_july, temperature_C ~ year4 + daynum + depth)

lm_final <- step(lm1)

print(lm_final$coefficients) #full model is the best model


#10
lm1 <- lm(data = ntl_july, temperature_C ~ year4 + daynum + depth)
print(summary(lm1))

```

11. What is the final set of explanatory variables that the AIC method suggests we use to predict temperature in our multiple regression? How much of the observed variance does this model explain? Is this an improvement over the model using only depth as the explanatory variable?

> Answer: 
The final output of the step() suggests that all three variables of "Year", 
"Day", and "Depth" are signficiant in determining the mean recorded temperature 
of lakes. The final model explains 74.11897% of the variability in mean recorded 
temperature which is slightly higher than that of the single-variable model. The 
single variable model does have a significantly higher AIC (53762) than that of
the fuller model (53674), meaning the fuller model is an improvement upon the 
single-variable.



---
## Analysis of Variance

12. Now we want to see whether the different lakes have, on average, different temperatures in the month of July. Run an ANOVA test to complete this analysis. (No need to test assumptions of normality or similar variances.) Create two sets of models: one expressed as an ANOVA models and another expressed as a linear model (as done in our lessons).

```{r anova.model}
#12
lake_anova <- aov(data = ntl_july, temperature_C ~ lakename)
print(summary(lake_anova))
lake_lm <- lm(data = ntl_july, temperature_C ~ lakename)
print(summary(lake_lm))

```

13. Is there a significant difference in mean temperature among the lakes? Report your findings. 

> Answer: According to both the ANOVA and linear models, a lake's given name has 
a significant effect in determining the mean recorded temperature. The ANOVA 
outputs a signficiant low p-value, meaning the null hypthesis (lake name has no 
significant effect on mean recorded temperature) is rejected. The linear model 
compares each lake to the "intercept" or a given lake's mean recorded 
temperature. It outputs significantly low p-values for each lakename as well, 
meaning that each lake has a significantly different mean temperature from the 
lake used as intercept.

14. Create a graph that depicts temperature by depth, with a separate color for each lake. Add a geom_smooth (method = "lm", se = FALSE) for each lake. Make your points 50 % transparent. Adjust your y axis limits to go from 0 to 35 degrees. Clean up your graph to make it pretty. 

```{r scatterplot.2}
#14.

ggplot(data = ntl_july, aes(y = temperature_C, x = depth,
                            color = lakename)) + 
  geom_point(alpha = 0.5, size = .5) +
  geom_smooth(method = "lm", se = F, size = 0.75) +
  ylim(c(0,35)) + 
  labs(x = "Depth(m)", y = "Temperature(C)", 
  title = "Mean July Lake Temp.\n vs.
       Depth Level", color = "Lake")

```

15. Use the Tukey's HSD test to determine which lakes have different means.

```{r tukey.test}
#15
TukeyHSD(lake_anova)
```

16.From the findings above, which lakes have the same mean temperature, statistically speaking, as Peter Lake? Does any lake have a mean temperature that is statistically distinct from all the other lakes?

>Answer: Crampton Lake & Central Long Lake, Ward Lake & Crampton Lake, 
Hummingbird Lake & East Long Lake, Tuesday Lake & East Long Lake, 
Tuesday Lake & Hummingbird Lake, West Long Lake & Hummingbird Lake, 
Peter Lake & Paul Lake, Ward Lake & Paul Lake, West Long Lake & Tuesday Lake are
pairs of lake with p-adjacent values above 0.05. Their mean recorded 
temperatures are not signficantly different.There is no one lake with a mean 
recorded temperature signficantly different from the others.

 

17. If we were just looking at Peter Lake and Paul Lake. What's another test we might explore to see whether they have distinct mean temperatures? 

>Answer: A two sample t test to test whether the mean recorded temperatures of 
Peter and Paul Lake are significantly different. The hypotheseses would be: 
H0: mean_temp_Paul - mean_temp_Peter == 0 / 
Ha: mean_temp_Paul - mean_temp_Peter != 0.


