---
title: "Assignment 09: Data Scraping"
author: "Curtis Cha"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

# Total points:

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on data scraping. 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your last name into the file name (e.g., "Fay_09_Data_Scraping.Rmd") prior to submission.


## Set up 
1. Set up your session:

* Check your working directory
* Load the packages `tidyverse`, `rvest`, and any others you end up using.
* Set your ggplot theme

```{r, message = FALSE}
#1
library(pacman)
pacman::p_load(tidyverse, rvest, lubridate, ggplot2, cowplot)

```

2. We will be scraping data from the NC DEQs Local Water Supply Planning website, specifically the Durham's 2019 Municipal Local Water Supply Plan (LWSP): 
 * Navigate to https://www.ncwater.org/WUDC/app/LWSP/search.php
 * Change the date from 2020 to 2019 in the upper right corner.
 * Scroll down and select the LWSP link next to Durham Municipality. 
 * Note the web address: <https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2020>

Indicate this website as the as the URL to be scraped. (In other words, read the contents into an
`rvest` webpage object.)

```{r set.the.scraping.website}
#2

website <- read_html(
  "https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2020")
```

3. The data we want to collect are listed below:

* From the "1. System Information" section:
 * Water system name
 * PSWID
 * Ownership
 
* From the "3. Water Supply Sources" section:
 * Average Daily Use (MGD) - for each month

In the code chunk below scrape these values, assigning them to three separate variables.

>HINT: The first value should be "Durham", the second "03-32-010", the third "Municipality", and the last should be a vector of 12 numeric values, with the first value being 36.0100.

```{r scrape.the.data}
#3
water.system.name <- 'div+ table tr:nth-child(1) td:nth-child(2)'
pswid <- "td tr:nth-child(1) td:nth-child(5)"
ownership <- "div+ table tr:nth-child(2) td:nth-child(4)"
max.withdrawals.mgd <- "th~ td+ td"

```


4. Convert your scraped data into a dataframe. This dataframe should have a column for each of the 4 variables scraped and a row for the month corresponding to the withdrawal data. Also add a Date column that includes your month and year in data format. (Feel free to add a Year column too, if you wish.)

>TIP: Use `rep()` to repeat a value when creating a dataframe.

>NOTE: It's likely you won't be able to scrape the monthly widthrawal data in order. You can overcome this by creating a month column in the same order the data are scraped: Jan, May, Sept, Feb, etc...

5. Plot the max daily withdrawals across the months for 2020

```{r create.a.dataframe.from.scraped.data}
#4

water_system <- website %>% 
  html_nodes(water.system.name) %>% html_text()
ID <- website %>%  
    html_nodes(pswid) %>%  html_text()
Ownership <- website %>% 
   html_nodes(ownership) %>% html_text()
max_withdrawals <- website %>%
    html_nodes(max.withdrawals.mgd) %>% html_text()

months <- c("Jan", "May", "Sep", "Feb", "Jun", 
            "Oct", "Mar", "Jul", "Nov", "Apr", "Aug", "Dec")
  
df <- data.frame("Water System" = rep(water_system, 12),
                 "PSWID" = rep(ID,12),
                 "Ownership" = rep(Ownership,12),
                 "Month" = my(paste(months,"-2020", sep ='')),
                 "Max_Withdrawals_mgd" = as.numeric(max_withdrawals))
  
df

#5

durham_2020 <- ggplot(df, aes(x = Month, y = Max_Withdrawals_mgd)) +
  geom_point() +
  labs(title = "2020 Monthly Max Withdrawls - Durham", 
       x = "Months", 
       y = "Max Withdrawal per Day (MGD)")

durham_2020

```

6. Note that the PWSID and the year appear in the web address for the page we scraped. Construct a function using your code above that can scrape data for any PWSID and year for which the NC DEQ has data. **Be sure to modify the code to reflect the year and site scraped**.

```{r construct.a.scraping.function}
#6.

scrape.it <- function(pwsid, year) {
  
  #website 
  website <- read_html(paste0(
    'https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=', 
                                  pwsid, '&year=', year))
  
  #html_elemnts
  water.system.name <- 'div+ table tr:nth-child(1) td:nth-child(2)'
  pswid <- "td tr:nth-child(1) td:nth-child(5)"
  ownership <- "div+ table tr:nth-child(2) td:nth-child(4)"
  max.withdrawals.mgd <- "th~ td+ td"

  water_system <- website %>% html_nodes(water.system.name) %>% html_text()
  ID <- website %>% html_nodes(pswid) %>%  html_text()
  Ownership <- website %>% html_nodes(ownership) %>% html_text()
  max_withdrawals <- website %>% html_nodes(max.withdrawals.mgd) %>% html_text()
  
  #dataframe
  months <- c("Jan", "May", "Sep", "Feb", "Jun", 
              "Oct", "Mar", "Jul", "Nov", "Apr", "Aug", "Dec")
  
  df <- data.frame("Water System" = rep(water_system, 12),
                   "PSWID" = rep(ID,12),
                   "Ownership" = rep(Ownership,12),
                   "Month" = my(paste(months,"-", year, sep ='')),
                   "Max_Withdrawals_mgd" = as.numeric(max_withdrawals))

  return(df)
}

```

7. Use the function above to extract and plot max daily withdrawals for Durham (PWSID='03-32-010') for each month in 2015

```{r fetch.and.plot.Durham.2015.data}
#7
durham_2015 <- scrape.it('03-32-010', '2015')
durham_2015

```

8. Use the function above to extract data for Asheville (PWSID = 01-11-010) in 2015. Combine this data with the Durham data collected above and create a plot that compares the Asheville to Durham's water withdrawals.

```{r fetch.and.plot.Asheville.2015.data}
#8
ashe_2015 <-scrape.it('01-11-010', '2015')

ashe_durham_2015 <- rbind(ashe_2015, durham_2015)

ggplot(ashe_durham_2015, aes(x = Month, 
                             y = Max_Withdrawals_mgd, 
                             color = Water.System)) + 
  geom_point() + 
  labs(title = "Water Usage (2015 Asheville)",
       y = "Max Withdrawals (MGD)")
```


9. Use the code & function you created above to plot Asheville's max daily withdrawal by months for the years 2010 thru 2019.Add a smoothed line to the plot.

```{r}
#9

scrape.it <- function(pwsid, year) {
  
  #website 
  website <- read_html(paste0(
    'https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=', 
                                  pwsid, '&year=', year))
  
  #html_elemnts
  water.system.name <- 'div+ table tr:nth-child(1) td:nth-child(2)'
  pswid <- "td tr:nth-child(1) td:nth-child(5)"
  ownership <- "div+ table tr:nth-child(2) td:nth-child(4)"
  max.withdrawals.mgd <- "th~ td+ td"

  water_system <- website %>% html_nodes(water.system.name) %>% html_text()
  ID <- website %>% html_nodes(pswid) %>%  html_text()
  Ownership <- website %>% html_nodes(ownership) %>% html_text()
  max_withdrawals <- website %>% html_nodes(max.withdrawals.mgd) %>% html_text()
  
  #dataframe
  months <- c("Jan", "May", "Sep", "Feb", "Jun", "Oct", 
              "Mar", "Jul", "Nov", "Apr", "Aug", "Dec")
  
  df <- data.frame("Water System" = rep(water_system, 12),
                   "PSWID" = rep(ID,12),
                   "Ownership" = rep(Ownership,12),
                   "Month" = my(paste(months,"-", year, sep ='')),
                   "Max_Withdrawals_mgd" = as.numeric(max_withdrawals))
  return(df)
}

years <- c(2010, 2011, 2012, 2013, 2014, 2015, 2016, 2017, 2018, 2019)
df_fin <- data.frame()

for (i in years) {
  df <- scrape.it('01-11-010', i)
  ifelse(i == 2010, 
         df_fin <- df, 
         df_fin <- rbind(df_fin, df)
  )
  }

ggplot(df_fin, aes(x = Month, y = Max_Withdrawals_mgd)) + 
  geom_point() + 
  geom_smooth(method = lm, se = F) +
  labs(title = "Water Usage (2010-2019 Asheville)",
       y = "Max Withdrawals (MGD)")

```

>Question: Just by looking at the plot (i.e. not running statistics), does Asheville have a trend in water usage over time?

Asheville's water usage (based on Monthly Max Withdrawals) has increased over time.